AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference

### 2019-08-20 paper/learning/notes.tex
added plots - still to update text

There are two different forms of rejection samplers found with in simulators and program code. 
In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 
sample from, but can evaluate.
We then define a proposal using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from.
Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
In the standard statistical setting, see algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
Here $M$ is a constant that has a finite bound, $M \in (1, \infty)$ and the likelihood ratio $\frac{f(x)}{g(x)}$ is defined over the 
support of $X$. 
$M$ must also satisfy the following inequality $f(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $f(x)>0$. 
We construct this acceptance probability by sampling from a uniform distribution defined on the interval 
$(0,1)$, which the proof ultimately relies on to ensure that a sample generated from the rejection sampler directly 
corresponds to a sample from the true, normalized, target density $f(x)$. 
But, provided we can calculate $P_{\text{accept}}(\mathcal{D} \leq \mathcal{C})$ for the given density, or mass, function
 $\mathcal{D}$ and condition $\mathcal{C}$, we can also satisfy the standard proof with a little more work.
 However, as we will demonstrate, rejection samplers 
 need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
 More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
\begin{algorithm}
  \label{alg:rsstats}
  \small
   \caption{ABC: Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \label{alg:rssim}
  \small
   \caption{Simulator: Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim p(x)$ 
       \State $\mathbb{I}(g(x) > 0)$
       \State $ Y \sim p(X=x | \mathbb{I}(g(x) > 0))$
       \State accept $X_i$ with $p(\mathbb{I}(g(x) > 0))$
\end{algorithmic}
\end{algorithm}
Thus, the probability of generating a sample $X = x$ from the true distribution can be defined, using Bayes rule, 
as: 
\begin{equation}
  p(X=x| \mathbb{I}) = \frac{p(\mathbb{I} | X=x)p(X=x)}{p(\mathbb{I})} 
\end{equation}
where $p(\mathbb{I} | X=x) = \{0,1\}$.

@@

If we have direct access to a rejection sampling loop then it is indeed 
possible to construct the density, up to a constant of proportionality. 
That is:
\begin{align*}
z \sim g(\theta) \\
do something \\
\alpha = p_{accept}(z,\theta) \\
\mathtt{if} Bern(\alpha) = 1: \\
return z
\end{align*}
thus the density of $y,\theta \propto q(\theta) \alpha$.
However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template such as R1 and R2. 
\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
extract all variables, other than $\alpha$ in an automated way.
However, without being able to easily extract $\alpha$ this means that we are unable to 
correctly build the form of the density and thus are limited in our ability to leverage 
powerful inference methods such as MCMC and Variational inference.
Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
This enables us to learn functions that we can easily sample from and via the amortization process construct the density of the entire
simulator.

The simulated densities for the rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
\begin{figure}
  \centering
  \includegraphics[width=0.2\textwidth, height=0.2\textheight, keepaspectratio]{rs_plots.png}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange}.}
  \label{fig:rs_simulated}
\end{figure}

@@

\paragraph{observations} As the functions are relatively simple the loss functions it gets t
to a loss of around 0.6 and stops decreasing. Although, this seems low, I don't think that it is
actually good, because that would mean on average, for each data point there is a loss of $\pm$ 0.6
which per data point is quite bad. That may be for a number of reasons, using MSE loss, using a simple
model architecture. It really struggles to model the tails. It may require a much larger batch input size. 
. Increasing the batch size helped to correct for the peaked density, however the behavior is still odd.
The peaked behaviour still does not disappear, this, I believe is due to the loss function averaging 
over all points.

>>>>

There are two different forms of rejection samplers found with in simulators and program code. 
In both instances we we have $X$, a random variable whose density is $f(x)$ that we cannot easily 
sample from, but can evaluate.
We then define a proposal using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from.
Where the two methods diverge is in how the acceptance probability is defined and how the samples are generated.
In the standard statistical setting, see Algorithm~\ref{alg:rsstats}, we generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
Here $M$ is a constant that has a finite bound, $M \in (1, \infty)$ and the likelihood ratio $\frac{f(x)}{g(x)}$ is defined over the 
support of $X$. 
$M$ must also satisfy the following inequality $f(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $f(x)>0$. 
We construct this acceptance probability by sampling from a uniform distribution defined on the interval 
$(0,1)$, which the proof ultimately relies on to ensure that a sample generated from the rejection sampler directly 
corresponds to a sample from the true, normalized, target density $f(x)$. 
But, provided we can calculate $P_{\text{accept}}(\mathcal{D} \leq \mathcal{C})$ for the given density, or mass, function
 $\mathcal{D}$ and condition $\mathcal{C}$, we can also satisfy the standard proof with a little more work.
 However, as we will demonstrate, rejection samplers 
 need not take this form and arise in many nested structures within probabilistic programs~\cite{rainforth2018nesting} and simulators~\cite{smith2008towards,bershteyn2018implementation,gleisberg2009event}.
 More generally speaking we can write a rejection sampler in the form of Algorithm~\ref{alg:rssim}
\begin{algorithm}
  \small
   \caption{ABC: Rejection Sampling}
   \label{alg:rsstats}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \small
   \caption{Simulator: Rejection Sampling}
   \label{alg:rssim}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim p(x)$ 
       \State $\mathbb{I}(g(x) > 0)$
       \State $ Y \sim p(X=x | \mathbb{I}(g(x) > 0))$
       \State accept $X_i$ with $p(\mathbb{I}(g(x) > 0))$
\end{algorithmic}
\end{algorithm}
Thus, the probability of generating a sample $X = x$ from the true distribution can be defined, using Bayes rule, 
as: 
\begin{equation}
  p(X=x| \mathbb{I}) = \frac{p(\mathbb{I} | X=x)p(X=x)}{p(\mathbb{I})} 
\end{equation}
where $p(\mathbb{I} | X=x) = \{0,1\}$.

@@

If we have direct access to a rejection sampling loop then it is indeed 
possible to construct the density, up to a constant of proportionality. 
That is:
\begin{align*}
z \sim g(\theta) \\
do something \\
\alpha = p_{accept}(z,\theta) \\
\mathtt{if} Bern(\alpha) = 1: \\
return z
\end{align*}
thus the density of $y,\theta \propto q(\theta) \alpha$.
However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
For any rejection sampling block, written in a generic code base, not in the form of R3, this is an incredibly hard 
task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template. 
\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}.
y leveraging the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
extract all variables, other than $\alpha$ in an automated way.
However, without being able to easily extract $\alpha$ this means that we are unable to 
correctly build the form of the density and thus are limited in our ability to leverage 
powerful inference methods such as MCMC and Variational inference.
Thus, we propose another approach, automating the amortization of rejection sampling loops of type R1, R2 and R3. 
This enables us to learn functions that we can easily sample from and via the amortization process construct the density of the entire
simulator.

The simulated densities for the rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
\begin{figure}
  \label{fig:rs_simulated}
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth, height=0.6\textheight, keepaspectratio]{rs_plots.png}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth, height=0.6\textheight, keepaspectratio]{rs3_plot.png}
  \end{minipage}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange} and rejection sampling block 3, \emph{red}.}
 \end{figure}

@@

\paragraph{observations} 
Using a standard $L2$-loss function $L_{2} = (y_{pred} - y_{true})^{2}$
As the functions are relatively simple the loss functions it gets t
to a loss of around 0.6 and stops decreasing. Although, this seems low, I don't think that it is
actually good, because that would mean on average, for each data point there is a loss of $\pm$ 0.6
which per data point is quite bad. That may be for a number of reasons, using MSE loss, using a simple
model architecture. It really struggles to model the tails. It may require a much larger batch input size. 
. Increasing the batch size helped to correct for the peaked density, however the behavior is still odd.
The peaked behaviour still does not disappear, this, I believe is due to the loss function averaging 
over all points.
